```{r setup} #setup is keyword. this chunk runs before others}
#install.packages('dplyr')
library(dplyr)
library(ggplot2)
```

Practicing `dplyr` functions using `starwars` dataset

```{r}
starwars
head(starwars)
?starwars
```

## Subset rows with `filter`

`filter()` allows us to subset the **rows** of a dataset based on the values

```{r}
## filter for rows with human Star Wars characters
filter(starwars, species == 'Human')

## save to new R object
humans <- filter(starwars, species == 'Human')
humans
filter(humans, sex == 'male')

## filter multiple columns at once
filter(starwars, species == 'Human', sex == 'male')
```

We can use different comparison operators to filter:

-   `!` is "not"

-   `==` is "equal"

-   `&` is "and"

-   `|` is "or"

-   `<` is "less than"

-   `>` is "greater than"

-   `is.na()` returns `TRUE` if value is `NA`

```{r}
## return characters with a missing mass
filter(starwars, is.na(mass))
```

```{r}
## get all characters that are NOT human
filter(starwars, species != 'human')
```

variable_name `%in%` vector_of_values returns TRUE if the value of variable_name is equal to one of the values in vector_of_values. Shortcut for variable_name == vector_of_values[1] \| variable_name == vector_of_values[2] ...

```{r}
skin_color <- 'grey'
skin_color %in% c('yellow','fair','tan') # returns false because 'grey' not in vector of colors

## filter for characters with brown or blond hair
starwars
filter(starwars, hair_color %in% c('brown', 'blond'))
```

grepl("characters_of_interest", variable_name) returns TRUE if the value of variable_name contains the characters_of_interest.

```{r}
filter(starwars, grepl('blond', hair_color))

## get all characters with blue in their eyes
filter(starwars, grepl('blue', eye_color))
```

Use `<`, `>`, `<=`, `>=` to compare numeric variables or categorical variables

-   nominal categorical variables (character vectors) are ranked alphabetically

-   ordinal categorical variables (factors) are ranked based on built-in order

```{r}
## starwars characters with name starting with letters before "D"
filter(starwars, name < 'D') #ranking based on alphabetical order

str(diamonds)
filter(diamonds, cut > "Good") #ranking based on levels of the ordered factor cut
"d" < "D" #true, lowercase < uppercase
```

between(x, left, right) is shorthand for x \>= left & x \<= right

```{r}
between(5,1,7) #true because 1<=5<=7
filter(starwars, between(mass, 50, 100))
```

## Practice

Find all starwars characters that are:

1.  missing hair color
2.  shorter than 100cm and are not droids
3.  born between years 100 and 200
4.  another of your choice

```{r practice_filters}
head(starwars)
filter(starwars, is.na(hair_color))
filter(starwars, height<100, species!='Droid')
filter(starwars, between(birth_year,100,200))
filter(starwars, grepl('gray',eye_color))
filter(starwars, starships != "character(0)")
```

## Order rows with `arrange`

`arrange()` reorders rows based on the values of the column(s)

```{r}
## order by character names
arrange(starwars, name)
#order: aAbBcC (a<A<b...)
#NA always at end

## order by species and birth year
arrange(starwars, species, birth_year) #birth years ordered within species groups

## desc() to get birth years in descending order
arrange(starwars, species, desc(birth_year))
```

## Practice

Arrange starwars dataset:

1.  from tallest to shortest
2.  shortest to tallest by species
3.  alphabetize characters by their hair color then eye color

```{r practice_arrange}
arrange(starwars, desc(height))
arrange(starwars, height, species) #order by height, tiebreaker is alphabetic species
arrange(starwars, species, height) #order by species, within species order shortest to tallest
arrange(starwars, hair_color, eye_color)
```

## Subset columns with `select()`

```{r}
select(starwars, hair_color, skin_color, eye_color)
character_colors <- select(starwars, hair_color, skin_color, eye_color)
character_colors
```

We can also select multiple columns in a row

```{r}
select(starwars, name:eye_color)
```

We can also remove columns by name using a -

```{r}
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles))
select(starwars, -c(films:starships))
```

Helpful functions:

`everything()` is a shortcut for "everything else"

```{r}
select(starwars, species, everything()) #moved species to front, then added everything else

## alphabetize columns
starwars_names <- names(starwars)
select(starwars, starwars_names)
```

`starts_with("abc")` matches column names that start with "abc"

`ends_with("abc")` ... end with "abc"

`contains("abc")` ... contain "abc"

```{r}
select(starwars, starts_with("H"))
select(starwars, contains("or"))
```

Practice

1.  select each character's homeworld and species. what question might this subset of variables help answer?
2.  brainstorm as many ways as possible to select hair_color, skin_color, eye_color, and birth_year

```{r}
head(starwars)
select(starwars, homeworld, species) #might help answer which species belong to each homeworld
select(starwars, hair_color, skin_color, eye_color, birth_year)
select(starwars, contains("color"), birth_year)
select(starwars, contains("_"))
select(starwars, hair_color:birth_year)
```

## Add new variables with `mutate`

```{r}
starwars_small <- select(starwars, height, mass, species)
mutate(starwars_small,
       height_m = height/100
       ) #calculates row-wise
```

Create multiple variables at once and we can refer to variables we just created

```{r}
mutate(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)

## only keep new variables
transmute(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)
```

Summary stats give one value for entire column: sum(), mean()

```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm=T))
mutate(starwars_small,
       row_addition = height + mass,
       summary_sum = sum(height, mass, na.rm=T))
```

Group by var using .by parameter

```{r}
mutate(starwars_small,
       species_mass = mean(mass, na.rm=T),
       .by = species)
```

Change the type of a variable

```{r}
mutate(starwars_small,
       mass = as.character(mass))
mutate(starwars_small,
       mass = as.logical(mass)) # if number is >0, evaluates to TRUE
```

Practice

1.  convert mass to pounds (1 kg \~ 2.2 lbs)

```{r}
mutate(starwars_small, mass_lbs = mass*2.2)
```

## Summarize data with `summarize`

```{r}
summarize(starwars, 
          mean_mass = mean(mass, na.rm=T))

#group by species
summarize(starwars,
          mean_mass = mean(mass, na.rm=T),
          .by = species)
```

The only columns kept are the columns you group by and the columns you create.

n() is a helpful function that returns the number of rows in a group

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm=T),
          count = n(),
          .by = species)

##group by multiple groups
summarize(starwars,
          mean_mass = mean(mass, na.rm=T),
          count = n(),
          .by = c(species, sex))
```

## Renaming

rename(data, new_name = old_name)

```{r}
new_starwars <- rename(starwars,
       birthYear = birth_year)
new_starwars
```
